---
title: "CONSTRUCTING DYADIC POSTHOC DROP-MODELS - Aura Raulo 2022"
output: html_notebook
---

When we have constructed a dyadic regression model predicting microbiome similarity with different pairwise effects, we can use a dropmodel-approach to deconstruct these whole-microbiome-level trends into the level of individual bacterial taxa. This allows us to perform a sort of dyadic indicator analysis, and ask post hoc questions such as which microbes are driving specific microbiome-similarity effects. It allows us to decompress the overall microbiome variation explained by various predictors into its components: Variation in the sharing of which bacterial taxa is explained by which predictor the most. 

Here we decompress the signals of different transmission effects on overall microbiome sharing into the level of distinct microbial taxa (genera) by asking "If social association (social transmission effect) and spatial overlap (environmental transmission effect) among pairs of mice independently predict their overall level of microbiome sharing (Jaccard microbiome similarity = proportion of taxa shared), then which bacteria are responsible for the social vs environmental transmission effect? In other words, when there is evidence for higher microbiome similarity among more socially AND spatially close mouse-pairs, is this due to sharing of the same or different subset of the microbiome, i.e. which bacterial taxa are transmitted socially and which environmentally? 

Note: In this example data set the spatial signal on microbiome similarity is actually not significant (credible intervals overlap zero), but we will use both social and spatial effects as an example of whole-microbiome-level effect-deconstruction, since we know that in a larger data set (of which this is a subset) both effects are significant drivers of microbiome similarity (See Raulo et al., 2021).

This modeling framework relies on measuring each microbial taxon's importance in each whole-microbiome-level trend. The whole-microbiome-level transmission signals are deconstructed to taxon-level by evaluating the extent to which dropping each taxon from the data increases the uncertainty around social/spatial transmission effect. This can be viewed as a score of importance that captures each taxon’s influence over the detectability of a given overall transmission signal. 

Finally, by calculating each taxon’s relative importance for social vs. environmental (spatial) transmission signal, we can estimate their reliance on a given transmission mode over the other. 

We will start by making a loop that derives the dyadic data table over and over again, making different versions of Jaccard microbiome similarity, each time dropping one microbial genus from the data. We will start by dropping genus "none" to get the baseline effects to which we can compare the posterior distributions of the transmission effects after each drop.

1. Prepare data
```{r}
library(phyloseq)
library(MCMCglmm)

#Read in microbiome data and associated sample data in phyloseq format. These data sets contain either 70 samples of 70 individuals
micdata<-readRDS( "Mouse_microbiome_idwise.rds")
#make a key for the order of sample names and their associated individual IDs.
key<-data.frame(ID=sample_data(micdata)$ID, Sample_name=sample_data(micdata)$Sample_name)
#Read in dyadic data frame
data.dyad_REAL<-readRDS("dyadic_model_data.rds")

# Data wrangling step: Give all unclassified genera a name based on their family or order
tax<-as.data.frame(tax_table(micdata))
tax[which(tax$Genus=="g__"),]$Genus<-paste0("Unknown_genus_in_",tax[which(tax$Genus=="g__"),]$Family)
tax[which(is.na(tax$Genus)),]$Genus<-paste0("Unknown_genus_in_",tax[which(is.na(tax$Genus)),]$Family)
tax[which(tax$Genus=="Unknown_genus_in_f__"),]$Genus<-paste0("Unknown_genus_in_",tax[which(tax$Genus=="Unknown_genus_in_f__"),]$Order)

#Make a list of the bacterial genera present in the data and how many unique taxa (ASVs) each genus has
tax_G<-as.data.frame(table(tax[,6]))
genuses<-tax_G$Var1
genuses<-c("none",genuses)

```
 Instead of brms, we wish to use MCMCglmm with gaussian error structure as the root model, since it is faster than brms and often gives similar estimates than brms-betaregression when the proportional response variable is non-skewed like here. Before we write the drop model loop, lets test that MCMCglmm model with gaussian error structure is comparable to our brms betaregression model
```{r}
#brms betaregression
brms_model<-readRDS("model1.rds")
summary(brms_model)

#MCMCglmm gaussian model
mcmcglmm_model<-MCMCglmm(Microbiome_similarity~1+age_similarity+ sex_combination+ spatial_distance+Temporal_distance+Social_proximity, 
             data=data.dyad_REAL,
             family= "gaussian",
             random =~ mm(IDA+IDB), 
             verbose=FALSE)

summary(mcmcglmm_model)


```
Drop-model loop
Recommended to run across 10-40 cores
One MCMCglmm model took 10 min on macbook pro 2022

```{r}
#Estimate run time
dropnumber<-length(genuses)#323
model_minutes<-10
cores<-40
runtime<-(dropnumber*model_minutes)/cores

paste("across", cores, "cores,", "mcmcglmm-dropmodel loop with" ,dropnumber,"genuses will take approx." ,as.character(runtime),"minutes to run")

#Parallelize dropmodels across cores
library(parallel)
library(doParallel)

  #start cluster
dropRes_list<-list()
cl <- parallel::makeCluster(40, type="FORK") 

doParallel::registerDoParallel(cl)

  dropRes_list<-foreach(i = 1:length(genuses)) %dopar% { 
    library(phyloseq)
    #choose the genus to drop and prune the taxa to keep everything else
    gen.i<-genuses[i]
    taxa_tokeep<-rownames(tax[which(tax$Genus!=genuses[i]),])
    mic.i<-prune_taxa(taxa_tokeep, micdata)
    #Calculate Jaccard and Bray-Curtis microbiome dissimilarity for each mouse pair
    JACM.i<- as.matrix(phyloseq::distance(mic.i, method="jaccard"))
    BRAY.i<- as.matrix(phyloseq::distance(mic.i, method="bray"))
    #Unravel dissimilarity matrices into vectors
    bray<-c(as.dist(BRAY.i))
    jac<-c(as.dist(JACM.i))
    
    #Make a new dyadic data frame from these vectors and order it to be in the same order as       the original dyadic data frame
    data.dyad.i<-data.frame(Jaccard=jac,BrayCurtis=bray)
    # extracting Sample_name-combinations of the matrix
    list<-expand.grid(key$Sample_name,key$Sample_name) 
    # This created sample-to-same-sample pairs as well. Get rid of these:
    list<-list[which(list$Var1!=list$Var2),] 
    # the resulting list still has both quantiles of the original matrix (i.e. all values are     doubled) in--> add 'unique' key and subset to one quantile only
    list$key <- apply(list, 1, function(x)paste(sort(x), collapse='')) 
    list<-subset(list, !duplicated(list$key)) 
    # Sample_name combinations are now in the same order as the lower quantile value vector
    # So we can add dyad name and each participant ID to dyadic dataframe
    data.dyad.i$Sample_A<-list$Var2
    data.dyad.i$Sample_B<-list$Var1
  
    # extracting combinations of individual IDs for each pair
    keyA<-key[,c("ID","Sample_name")]
    colnames(keyA)<-c("IDA","Sample_A")
    keyB<-key[,c("ID","Sample_name")]
    colnames(keyB)<-c("IDB","Sample_B")
  
    keyA<-keyA[match(data.dyad.i$Sample_A,keyA$Sample_A),]
    keyB<-keyB[match(data.dyad.i$Sample_B,keyB$Sample_B),]
  
    data.dyad.i$IDA<-keyA$IDA
    data.dyad.i$IDB<-keyB$IDB
  
    # Make sure you have no self comparisons in the data (This is the case by default here,       since we are using just one sample per individual)
    
    data.dyad.i<-data.dyad.i[which(data.dyad.i$IDA!=data.dyad.i$IDB),] #2415
  
    ### Combine new Jaccard variable with rest of dyadic data columns
    data.dyad<-data.dyad_REAL
    data.dyad$Jaccard<-data.dyad.i$Jaccard
    
    #factorize terms used for multimembership random structure and make sure levels are     same and in same order
    data.dyad$IDA<-as.factor(data.dyad$IDA)
    data.dyad$IDB<-as.factor(data.dyad$IDB)
    all(levels(data.dyad$IDA)==levels(data.dyad$IDB))#T

  # Scale all predictors not between 0-1 already to be between 0-1
    scalecols<-c("Temporal_distance" ,"spatial_distance")
    
    range.use <- function(x,min.use,max.use){ (x -   
    min(x,na.rm=T))/(max(x,na.rm=T)-min(x,na.rm=T)) * (max.use - min.use) + min.use }
  
    for(i in 1:ncol(data.dyad[,which(colnames(data.dyad)%in%scalecols)])){
      data.dyad[,which(colnames(data.dyad)%in%scalecols)][,i]<-range.use(data.dyad[,which(colnames(data.dyad)%in%scalecols)][,i],0,1)
      }
    #Transpose Jaccard dissimilarity to similarity
    data.dyad$Jaccard<-1-data.dyad$Jaccard
  
    #The MCMCglmm model
    dropmodel<-MCMCglmm(Microbiome_similarity~1+age_similarity+ sex_combination+ spatial_distance+Temporal_distance+Social_proximity, 
                      data=data.dyad,
                      family= "gaussian",
                      random =~ mm(IDA+IDB), 
                      verbose=FALSE)

    ASVs_dropped.i<-nrow(tax_table(micdata))-nrow(tax_table(mic.i))
    resdf.i<-data.frame(Genus_dropped=gen.i,
                      ASVs_dropped=ASVs_dropped.i,
                      Social_Estimate=summary(dropmodel)$solutions["Social_proximity",]["post.mean"],
                      Social_lCI=summary(dropmodel)$solutions["Social_proximity",]["l-95% CI"], 
                      Social_uCI=summary(dropmodel)$solutions["Social_proximity",]["u-95% CI"],
                      Spatial_Estimate=summary(dropmodel)$solutions["spatial_distance",]["post.mean"],
                      Spatial_lCI=summary(dropmodel)$solutions["spatial_distance",]["l-95% CI"], 
                      Spatial_uCI=summary(dropmodel)$solutions["spatial_distance",]["u-95% CI"]
                      )
    return(resdf.i)
  }
parallel::stopCluster(cl) 

saveRDS(dropRes_list,"dropRes_list.rds")

##rbind the resulting data frames to single master data frame
        dropResults<-data.frame(Genus_dropped=NA,
                                            ASVs_dropped=NA,
                                            Social_Estimate=NA,
                                            Social_lCI=NA,
                                            Social_uCI=NA,
                                            Spatial_Estimate=NA,
                                            Spatial_lCI=NA,
                                            Spatial_uCI=NA,
                                            habitat_Estimate=NA,
                                            habitat_lCI=NA,
                                            habitat_uCI=NA)

  for(j in 1:length(genuses)){
    dropResults<-rbind(dropResults,dropRes_list[[j]])
  }
dropResults<-dropResults[2:nrow(dropResults),]

dropResults$Genus_dropped2<-genuses

saveRDS(dropResults,"dropResults_genus.rds")

```
Calculate importance score for each bacterial genus over each transmission effect (Social association or spatial distance effect on microbiome sharing)

Here, the importance I(G) of microbial genus G on dyadic effect E (social of spatial effect on whole microbiome sharing) is defined as follows:

CIbr_increase(E_G)= CIbr(E_dropG)-CIbr(E_baseline),
If CIbr_increase(E_G)<0, CIbr_increase(E_G)=0,
I(G_E)= (CIbr_increase(E_G)/CIbr(E_baseline))/sqrt(no_ASVs),

where CIbr(E_baseline) is the breadth of credible intervals around effect E in the baseline model (with nothing dropped),

CIbr(E_dropG) is the breadth of credible intervals around effect E in the model where genus G is dropped,

CIbr_increase(E_G) is thus the amount of added uncertainty, that dropping genus G from the response variable will introduce around effect E. When credible interval breadth does not increase but in fact decreases after a drop (CIbr_increase(E_G)<0), we consider this genus to have no importnace over given effect, and thus we will set this to zero (CIbr_increase(E_G)=0).

I(G_E) is the importance genus G has on effect E, calculated as the increase in credible interval breadth (CIbr_increase) relative to the original credible interval breadth CIbr(E_baseline), divided by the square root of the number of unique sequence variants that were dropped as part of that genus, and multiplied by 100 to increase the scale. In other words it describes how much uncertainty is introduced around an effect per unit data lost. 
```{r}

#For each microbial genera, calculate their importance on social association effect on microbiome
Social_CIbr_baseline<-dropResults[which(dropResults$Genus=="none"),]$Social_CIbr
dropResults$Social_CIbr<-abs(dropResults$Social_uCI-dropResults$Social_lCI)
dropResults$Social_CIbr_increase<-dropResults$Social_CIbr-Social_CIbr_baseline
#Only those genera whose drop increases the uncertainty around an effect will be considered important for that effect:
dropResults[which(dropResults$Social_CIbr_increase<0),]$Social_CIbr_increase<-0
#Importance value is this increase in uncertainty divided by the square root of how many ASVs were dropped, and multiplied by 100 to increase the scale.
dropResults$IMPORTANCE_SOCIAL<-(dropResults$Social_CIbr_increase/Social_CIbr_baseline)/sqrt(dropResults$ASVs_dropped)*100

#For each microbial genera, calculate their importance on spatial distance effect on microbiome
Spatial_CIbr_baseline<-dropResults[which(dropResults$Genus=="none"),]$Spatial_CIbr
dropResults$Spatial_CIbr<-abs(dropResults$Spatial_uCI-dropResults$Spatial_lCI)
dropResults$Spatial_CIbr_increase<-dropResults$Spatial_CIbr-Spatial_CIbr_baseline
#Only those genera whose drop increases the uncertainty around an effect will be considered important for that effect:
dropResults[which(dropResults$Spatial_CIbr_increase<0),]$Spatial_CIbr_increase<-0
#Importance value is this increase in uncertainty divided by the square root of how many ASVs were dropped, and multiplied by 100 to increase the scale.
dropResults$IMPORTANCE_SPATIAL<-(dropResults$Spatial_CIbr_increase/Spatial_CIbr_baseline)/sqrt(dropResults$ASVs_dropped)*100

```
Now, microbes can have high importance for social, spatial, neither or both signals. To get a measure of which of the two transmission pathways each genus may rely on more, we will calculate a "reliance score" for microbial genus, defined as:

R(G_E1)= I(G_E1)*I(G_E1)/I(G_E2),

where R(G_E1) is the reliance of genus G on effect E1 (i.e. social transmission), calculated as the importance of genus G on effect E1 (I(G_E1)) weighted by the ratio of I(G_E1) and I(G_E2), i.e a measure of how much more importance they have for this effect E1 compared to the other effect E2.


```{r}
#For each microbial genera, calculate their reliance on social association effect over spatial effect on microbiome

dropResults$RELIANCE_SOCIAL<-dropResults$IMPORTANCE_SOCIAL*(dropResults$IMPORTANCE_SOCIAL/dropResults$IMPORTANCE_SPATIAL)

#For each microbial genera, calculate their reliance on spatial effect over social effect on microbiome

dropResults$RELIANCE_SPATIAL<-dropResults$IMPORTANCE_SPATIAL*(dropResults$IMPORTANCE_SPATIAL/dropResults$IMPORTANCE_SOCIAL)
```